A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 calculus.tex
restructuring

\begin{definition}
  A set~$C$ of transitions of an automaton~$\Automaton$ \emph{dominates}
  a  transition~$\Transition$, written
  $\Dominates(C, \Automaton, \Transition)$, if
  every accepting path $p$ of $\Automaton$
  with $\Transition \in p$ contains at least one transition from
  $C$. Notably, $\Dominates(\emptyset, \Automaton, \Transition)$ for
  every unreachable transition $\Transition$, and
  $\Dominates(\Set{\Transition}, \Automaton, \Transition)$ for every
  transition~$\Transition$.
\end{definition}

@@

We then introduce the two predicates into our calculus with the following
definitions:

\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$ holds for some
  automaton $\Automaton = \AutomatonTuple$, some map
  $\Map : \Alphabet^* \to \Monoid$ to a commutative monoid $\Monoid$,
  some transition selection function
  $\Filter:\: \Transitions \to \Naturals$, and some monoid
  element~$\MonoidElement \in \Monoid$ if $\MonoidElement$ is an
  element of the Parikh image of $\Language(\Automaton)$ modulo
  $\Map$, or more formally, when there is an accepting path
  $\Path = \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such
  that $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
  for all $t \in \Transitions$, and
  $\MonoidElement = \Map(\WordOf(\Path))$.
\end{definition}

@@

We present the rules of \Calculus{} for one automaton in
\cref{tbl:rules:single}. The rules operate on sets of formulas and
can be interpreted as rules of a one-sided sequent calculus, in
which all formulas are located in the antecedent~\cite{Fitting96a}.

Each rule should be read and applied bottom-up and relates
premises~$\SomeClause_1, \ldots, \SomeClause_k$ with some
conclusion~$\SomeClause$. When constructing a proof, we start with some
root~$\SomeClause$, and then apply proof rules to the goals of the proof until a
goal can be closed, or no more rule is applicable.

A proof in which all proof goals are closed witnesses that the formulas
in the root~$\SomeClause$ of the proof are inconsistent, and do not
have any solutions. A proof goal that cannot be closed, but to which no
more rules are applicable, gives rise to  a solution of the formulas in the
root~$\SomeClause$. Such an open proof goal will only contain formulas
in Presburger arithmetic allowing a solution to be computed using
standard algorithms~\cite{DBLP:books/daglib/0022394}.

We use the convention of splitting the formulas into linear (in)equalities
($\SomeInequalities$) and other formulas ($\SomeClause$), and assume that
predicates $\Image$ and $\Connected$ only occur positively. Since our rules
operate by adding and matching linear (in)equalities in a proof goal; we use the
shorthand of listing the matched inequalities as antecedents, e.g., in
\Propagate{}.

The filtering function~$\Filter$ is evaluated symbolically and can, in practice,
be read as a function from transitions to $\Naturals$-valued terms
(e.g.~\texttt{t} or~\texttt{t+1}). In our implementation \Catra{}, described in
\cref{sec:implementation}, $\Filter$~is a vector of fresh variables with the
same size as~$\Transitions$.

We use the shorthand notation~$\Transitions_\Automaton$ to refer to the
transitions of an automaton~$\Automaton$. Additionally, for an
automaton~$\Automaton = \AutomatonTuple$, we allow mapping the selection function
like so: $\Filter(\Automaton) = \Tuple{\States, \InitialState, \AcceptingStates,
\Set{\Transition \in \Transitions \SuchThat \Filter(\Transition) > 0}}$, i.e.
$\Automaton$~with only the transitions for which~$\Filter$ is positive. In this
instance, the basis for the matched linear inequalities is
implicitly~$\SomeInequalities$. Similarly, for our commutative monoid $\Monoid =
\left(X;\MonoidOp;0_{\Monoid}\right)$ and the map into it $\Map : \Strings
\rightarrow X$, we also allow mapping over transitions:
$\Map(\FromLabelTo{\State}{\Label}{\State'}) = \Map(\Label)$.

We require that rules can only be applied when they add new formulas
on every created branch (the notion of \emph{regularity} of a proof is
required~\cite{Fitting96a}). For example, this means that \Split{} can
only be applied to proof goals that contain neither
$\Filter(\Transition) = 0$ nor $\Filter(\Transition) > 0$, and can
never be applied to split on the same term twice on the same branch.
This suggests a proof strategy where you \Propagate{} when you can, \Split{}
when you must, and \Subsume{} when neither is possible anymore.

@@

  % SUBSUME
  \Subsume &
  \inferrule{\SomeInequalities,\SomeClause}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  \Split{} and \Propagate{}\newline cannot be applied \\
  \bottomrule
  \end{tabular}
  \caption{Derivation rules for one automaton.}\label{tbl:rules:single}
\end{table}
We use the symbolic function $\FlowEq(\Automaton, \Filter)$ that generates a set
of existentially quantified linear inequalities with the following definition,
where we assign fresh, existentially quantified variables to
$\FinalStateVar_\State, \Filter(\Transition)$ for every $\State \in
\AcceptingStates, \Transition \in \Transitions$:
\[
\begin{aligned}
  \FlowEq(\Automaton, \Filter) &= \sum\limits_{\State \in \AcceptingStates} \FinalStateVar_\State = 1 \land \AndComp{\State \in \States}{\In(\State, \Filter) - \Out(\State, \Filter)} = \Sink(\State) \land
  \AndComp{\State \in \AcceptingStates}{\FinalStateVar_\State \geq 0}\\
  \Sink(\State) & = 0 \text{ if } \State \not\in \AcceptingStates, \FinalStateVar_\State \text{ otherwise.} \\
  \In(\State, \Filter) & = \StartFlow(\State) + \sum_{\Transition \in \FromLabelTo{\State'}{}{\State}} \Filter(\Transition)\\
  \StartFlow(\State)  &= 1 \text{ if } \State = \InitialState, \text{ otherwise $0$.} \\
  \Out(\State, \Filter) &= \sum_{\Transition \in \FromLabelTo{\State}{}{\State'}} \Filter(\Transition)
\end{aligned}
\]

In addition to \cref{tbl:rules:single}, we assume the existence of a rule
\PresburgerClose{}, corresponding to a sound and complete solver for Presburger
formulae, and for the elements of~$\Monoid$.

The $\Propagate{}$ rule allows us to propagate (dis-)connectedness across
$\Automaton$. It states that we are only allowed to use transitions attached to
a reachable state, and is necessary to ensure connectedness in the presence of
cycles in~$\Automaton$.

\textsc{Expand} expands the predicate into its most basic rules; one set of
linear equations synchronising the transitions mentioned by~$\Filter$ to the
corresponding Monoid element~$\MonoidElement$, and the linear flow equations of
the standard Parikh image formulation, as described by~\FlowEq. Since
$\Connected$ and $\Image$ are partially redundant and the difference is covered
by~$\FlowEq$, we can remove the instance of~$\Image$ when applying~$\Expand$. In
this sense, we split the semantics of the $\Image$~predicate into its counting
aspect (covered by $\FlowEq$) and its connectedness aspect (covered by
$\Connected$).

Finally, $\Split{}$ allows us to branch the proof tree by trying to exclude a
contested transition from a potential solution before concluding that it must be
included. Intuitively, this is what guarantees our ability to make forward
progress by eliminating paths through~$\Automaton$.

A decision procedure for our predicate in a tableau-based automated theorem
prover would start by expanding the predicate using the $\Expand{}$~rule. A
theorem prover would perform algebraic substitution on the underlying constants
of~$\Filter$, boiling them down to choices of branches, which depend on one
single variable and loop transitions. This logic corresponds to the placement
of counters for optimally edge-profiling the CFG of a program, making up a
minimum-spanning tree of the automaton~\cite{path-profiling}.

In order to make the examples below tractable, we will assume the existence of a
rule \EquationReasoning{} that allows us to perform standard algebraic reasoning
on linear inequalities. This rule is not necessary for correctness or
completeness but shortens the examples considerably.

\subsection{An Example}\label{sec:single:example}
Here we will return to $\mathcal{A}'$ from \cref{sec:intuition} and performs the steps of
\cref{sec:intuition:algebra,sec:intuition:split} with the
formal calculus we just established, but exclude \cref{const:s2-in-b} and
therefore the entire automaton $\mathcal{B}$ since we introduce support for products of
automata in \cref{sec:multiple}. Note that this example then becomes
satisfiable, as $s_1 = \Char{dd}, i = 1, s_2 = \Char{adda}, l_A = 4$ is a
satisfying assignment to
\cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring}.

Starting with~$\mathcal{A}'$, $\Map$ extracts the increments of a transition, $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. = $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$. Initially, we let $\Filter$ map to fresh variables to obtain:
\begin{equation}\label{eq:single-filter}
  \begin{array}{rl@{\qquad}rl@{\qquad}rl}
    \Filter(\FromLabelTo{D}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A})
    &= \TransitionVar_1
    &
    \Filter(\FromLabelTo{F_A}{\Sigma/\begin{bmatrix} l_{A+} \end{bmatrix}}{F_A} )
    &= \TransitionVar_2   
    &
    \Filter(\FromLabelTo{I_A}{c/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A} ) 
    &=  \TransitionVar_3 
    \\
    \Filter(\FromLabelTo{I_A}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{D}) 
    &= \TransitionVar_4
    &
    \Filter(\FromLabelTo{I_A}{\Sigma/\begin{bmatrix} l_{A+} i_+\end{bmatrix}}{I_A}) 
    &= \TransitionVar_5  \\[3ex]
%
  \multicolumn{6}{c}{
    \FlowEq(\mathcal{A}', \Filter)  = \begin{aligned}
                       1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_3 + \TransitionVar_1\\
                    \TransitionVar_4 &= \TransitionVar_1 \\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &\geq \cancel{\TransitionVar_2}  \\
                    \end{aligned}}
  \end{array}
\end{equation}

Like in \cref{sec:a_1,sec:intuition:algebra}, we apply reasoning by equivalence
  to end up with fixed values for several of the transition variables
  $\TransitionVar_k$, and eventually we use \cref{const:more-inside-than-before}
  and \cref{const:something-before-and-after} to conclude that the transition
  directly from state $I_{A}$ to $F_{A}$ is incompatible with the constraint.

  Then, we remove the $\Connected{}$ predicate using \Subsume{} since no
   transition can become disconnected (and never could for $\mathcal{A}'$,
   since all looping transitions are attached to the initial and single
   accepting state and therefore always must be reachable from any accepting
   path). After this, the rest of the reasoning can be continued without any
   part of \Calculus{} since all values of the existentially quantified
   transition variables $\TransitionVar_1,\ldots,\TransitionVar_5$ compatible
   with
   \cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring,const:something-before-and-after}
   are represented by the flow equations introduced by $\Expand{}$. The
   derivation is shown in \cref{fig:derivation:single}.

\begin{figure}[ht]
  \centering
\begin{prooftree}
  \hypo[]{{\text{Regular solving finishes the proof}}}
  \infer1[\EquationReasoning{}: $i = 1$]{
  i = 1
  \land n = 2
  \land \TransitionVar_2 = l_A - 3
  \land \TransitionVar_3 = 0
  \land \TransitionVar_5 = 1
  \land \TransitionVar_1 = \TransitionVar_4 = 1
  }
  \infer1[]{
    \begin{aligned}
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2
    \land l_A - i - 2 > 0 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}}
  \infer1[\Subsume]{
    \begin{aligned}
    \Connected(\mathcal{A}', \Filter) \land 
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2 
    \land l_A - i - 2 > 0
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $\TransitionVar_1 = 1$ \cref{sec:a_1}]{
    \begin{aligned}
    \Connected(\mathcal{A}', \Filter) \land
    i = \TransitionVar_5 \land 
    n = 2\TransitionVar_1 
    \land n > i > 0 \\
    \land l_A - i - n > 0 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-elimination]{
    \begin{aligned}
    \Connected(\mathcal{A}', \Filter)  
    \land i = \TransitionVar_5 
    \land n = \TransitionVar_4 + \TransitionVar_1 
    \land n > i > 0
    \land l_A - i - n > 0
    \land \TransitionVar_4 = \TransitionVar_1 \\
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1
    \land 1 = \TransitionVar_3 + \TransitionVar_1 
    \land i > 0
    \land \TransitionVar_4 = \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand $\Sigma, \FlowEq$]{
    \begin{aligned}
    i = \TransitionVar_5 
    \land l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k 
    \land n > i > 0
    \land l_A - i - n > 0
    \land n = \TransitionVar_4 + \TransitionVar_1\\
    \land \Connected(\mathcal{A}', \Filter) 
    \land \FlowEq(\mathcal{A}', \Filter)
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \Connected(A', \Filter) \land
    \FlowEq(A', \Filter) \land
    \left[l_A, i, n\right] =
    \sum\limits_{\Transition \in \Transitions_{\mathcal{A}'} } \Map(\Transition) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{
    \Image{}_{\mathcal{A}', \Map}(\Filter, \left[l_A, i, n\right]) 
    \land n \geq i 
    \land l_A - i - n > 0
    \land i > 0
    }
\end{prooftree}
\caption{A derivation tree for an application of
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $\mathcal{A}'$ of \cref{fig:aut_a}. Note the liberal use of removal of equal
variables (e.g. if $x = y$ appears as a statement, we substitute $y$ for $x$
everywhere and drop $\ldots \land x=y \land \ldots$) to keep the size of the
equations manageable.}\label{fig:derivation:single}
\end{figure}

To obtain the automaton corresponding to the final model, one can evaluate
$\Filter(\Transition) = k$ for each transition $\Transition$ and construct a
path with each such transition taken $k$ times.

>>>>

@@

We introduce the two predicates that will be used by our calculus,
with the following definitions:
%
\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$ holds for some
  automaton $\Automaton = \AutomatonTuple$, some map
  $\Map : \Alphabet^* \to \Monoid$ to a commutative monoid $\Monoid$,
  some transition selection function
  $\Filter:\: \Transitions \to \Naturals$, and some monoid
  element~$\MonoidElement \in \Monoid$ if $\MonoidElement$ is an
  element of the Parikh image of $\Language(\Automaton)$ modulo
  $\Map$, or more formally, when there is an accepting path
  $\Path = \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such
  that $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
  for all $t \in \Transitions$, and
  $\MonoidElement = \Map(\WordOf(\Path))$.
\end{definition}

@@

@@

  % SUBSUME
  \Subsume &
  \inferrule{\SomeInequalities,\SomeClause}{\Connected(\Automaton, \Filter), \SomeInequalities, \SomeClause} &
  \Split{} and \Propagate{}\newline cannot be applied \\
  \bottomrule
  \end{tabular}
  \caption{Derivation rules for one automaton.}\label{tbl:rules:single}
\end{table}

The rules of \Calculus{} for one automaton are given in
\cref{tbl:rules:single}. The rules operate on sets of formulas and can
be interpreted as rules of a one-sided sequent calculus, in which all
formulas are located in the antecedent~\cite{Fitting96a}.
%
The rules relate premises~$\SomeClause_1, \ldots, \SomeClause_k$ with
some conclusion~$\SomeClause$. 
%
When constructing a proof, we start with some root~$\SomeClause$, and
then apply proof rules to the goals of the proof, in bottom-up
direction, until all goals are closed, or no more rule is applicable.

A proof in which all proof goals are closed shows that the formulas
in the root~$\SomeClause$ of the proof are inconsistent, and do not
have any solutions. A proof goal that cannot be closed, but to which no
more rules are applicable, gives rise to  a solution of the formulas in the
root~$\SomeClause$. Such an open proof goal will only contain formulas
in Presburger arithmetic allowing a solution to be computed using
standard algorithms~\cite{DBLP:books/daglib/0022394}.

We use the convention of splitting the formulas in proof goals into
linear (in)equalities ($\SomeInequalities$) and other formulas
($\SomeClause$), and assume that predicates $\Image$ and $\Connected$
only occur positively.
%Since our rules operate by adding and matching
%linear (in)equalities in a proof goal; we use the shorthand of listing
%the matched inequalities as antecedents, e.g., in \Propagate{}.
%
The transition selection function~$\Filter$ is represented
symbolically and can, in practice, be read as a function from
transitions to $\Naturals$-valued terms (e.g.~\texttt{t}
or~\texttt{t+1}). In our implementation \Catra{}, described in
\cref{sec:implementation}, $\Filter$~is a vector of fresh variables
with the same size as~$\Transitions$.

To ensure termination, rules can only be applied when they add new formulas
on every created branch (the notion of \emph{regularity} of a proof is
required~\cite{Fitting96a}). For example, this means that \Split{} can
only be applied to proof goals that contain neither
$\Filter(\Transition) = 0$ nor $\Filter(\Transition) > 0$, and can
never be applied to split on the same term twice on the same branch.

\medskip
%
The rule~\textsc{Expand} expands an $\SinglePredicateInstance$
predicate into the more basic predicate
$\Connected(\Automaton, \Filter)$, as well as linear equations
relating the transitions mentioned by~$\Filter$ with the monoid
element~$\MonoidElement$, and linear flow equations described
by~\FlowEq{} (below). Since $\Connected$ and $\Image$ are partially
redundant and the difference is covered by~$\FlowEq$, we can remove
the instance of~$\Image$ when applying~$\Expand$. In this sense, we
split the semantics of the $\Image$~predicate into its counting aspect
(covered by $\FlowEq$) and its connectedness aspect (covered by
$\Connected$).

In \textsc{Expand}, we use the shorthand
notation~$\Map(\FromLabelTo{\State}{\Label}{\State'}) = \Map(\Label)$,
i.e., we allow the homomorphism~$h$ to be applied also to
transitions~$\Transition$.
%
The predicate $\FlowEq(\Automaton, \Filter)$ represents the flow
equations to be generated when expanding
$\SinglePredicateInstance$. We assume that each application of the
predicate introduces fresh integer variables~$f_q$ for every accepting
state~$q \in \AcceptingStates$, and define:
%
\[
\begin{aligned}
  \FlowEq(\Automaton, \Filter) &~=~
   \AndComp{\State \in \AcceptingStates}{\FinalStateVar_\State \geq 0}
   \land
   \sum\limits_{\State \in \AcceptingStates} \FinalStateVar_\State = 1 \land \AndComp{\State \in \States}{\In(\State, \Filter) - \Out(\State, \Filter)} = \Sink(\State)\\
  \Sink(\State) & ~=~ 0 \text{ if } \State \not\in \AcceptingStates, \FinalStateVar_\State \text{ otherwise.} \\
  \In(\State, \Filter) & ~=~ \StartFlow(\State) + \sum_{\Transition \in\, (\FromLabelTo{}{}{\State})} \Filter(\Transition)\\
  \StartFlow(\State)  &~=~ 1 \text{ if } \State = \InitialState, \text{ otherwise $0$.} \\
  \Out(\State, \Filter) &~=~ \sum_{\Transition \in\, (\FromLabelTo{\State\,}{}{})} \Filter(\Transition)
\end{aligned}
\]

The rule~$\Split{}$ allows us to branch the proof tree by trying to exclude a
contested transition from a potential solution before concluding that it must be
included. Intuitively, this is what guarantees our ability to make forward
progress by eliminating paths through~$\Automaton$.

The $\Propagate{}$ rule allows us to propagate (dis-)connectedness
across $\Automaton$. It states that we are only allowed to use
transitions attached to a reachable state, and is necessary to ensure
connectedness in the presence of cycles in~$\Automaton$. The rule
makes use of the notion of \emph{dominating} sets of transitions, defined
as follows:
%
\begin{definition}
  A set~$C$ of transitions of an automaton~$\Automaton$ \emph{dominates}
  a  transition~$\Transition$, written
  $\Dominates(C, \Automaton, \Transition)$, if
  every accepting path $p$ of $\Automaton$
  with $\Transition \in p$ contains at least one transition from
  $C$. Notably, $\Dominates(\emptyset, \Automaton, \Transition)$ for
  every unreachable transition $\Transition$, and
  $\Dominates(\Set{\Transition}, \Automaton, \Transition)$ for every
  transition~$\Transition$.
\end{definition}

Finally, the rule~\Subsume{} can be applied when the connectedness of
an automaton has been ensured by exhaustive application of the other
rules. This suggests a proof strategy where you \Propagate{} when you
can, \Split{} when you must, and \Subsume{} when neither is possible
anymore.

In addition to \cref{tbl:rules:single}, we assume the existence of a rule
\PresburgerClose{}, corresponding to a sound and complete solver for Presburger
formulae, and for constraints over~$\Monoid$.

\medskip
%
A decision procedure would start from one or multiple
predicates~$\SinglePredicateInstance$ to be satisfied, possibly in
combination with other constraints about $m$. It would then first
expand the predicates using the $\Expand{}$~rule, and subsequently
apply the other rules to search for a solution.

As illustrated in \cref{sec:intuition}, a decision procedure can
also perform arithmetic rewriting of the occurring terms and
equations. Such reasoning is not necessary for correctness or
completeness, but it shortens the examples considerably; we
will therefore assume the existence of a
rule \EquationReasoning{} that allows us to perform standard algebraic reasoning
on linear arithmetic constraints.

%, boiling them down to choices of branches, which depend
%on one single variable and loop transitions. This logic corresponds to
%the placement of counters for optimally edge-profiling the CFG of a
%program, making up a minimum-spanning tree of the
%automaton~\cite{path-profiling}.

\subsection{An Example}\label{sec:single:example}
Here we will return to $\mathcal{A}'$ from \cref{sec:intuition} and perform the steps of
\cref{sec:intuition:algebra,sec:intuition:split} with the
formal calculus we just established, but exclude \cref{const:s2-in-b} and
therefore the entire automaton $\mathcal{B}$, since we introduce support for products of
automata in \cref{sec:multiple}. The example then becomes
satisfiable, as $s_1 = \Char{dd}, i = 1, s_2 = \Char{adda}, l_A = 4$ is a
satisfying assignment to
\cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring}.

Starting with~$\mathcal{A}'$, $\Map$ extracts the increments of a transition, $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$. Initially, we let $\Filter$ map to fresh variables to obtain, after some simplifications:
\begin{equation}\label{eq:single-filter}
  \begin{array}{rl@{\qquad}rl@{\qquad}rl}
    \Filter(\FromLabelTo{D}{d/[l_{A+}, n_+]}{F_A})
    &= \TransitionVar_1
    &
    \Filter(\FromLabelTo{F_A}{\Sigma/[ l_{A+} ]}{F_A} )
    &= \TransitionVar_2   
    &
    \Filter(\FromLabelTo{I_A}{c/[l_{A+}, n_+]}{F_A} ) 
    &=  \TransitionVar_3 
    \\
    \Filter(\FromLabelTo{I_A}{d/[l_{A+}, n_+]}{D}) 
    &= \TransitionVar_4
    &
    \Filter(\FromLabelTo{I_A}{\Sigma/[ l_{A+}, i_+]}{I_A}) 
    &= \TransitionVar_5  \\[3ex]
%
  \multicolumn{6}{c}{
    \FlowEq(\mathcal{A}', \Filter)  = \begin{aligned}
                       1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_4 + \TransitionVar_3\\
                    \TransitionVar_4 &= \TransitionVar_1 \\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &= \cancel{\TransitionVar_2} + 1  \\
                    \end{aligned}}
  \end{array}
\end{equation}

The proof tree is shown in \cref{fig:derivation:single}.  Like in
\cref{sec:a_1,sec:intuition:algebra}, by arithmetic reasoning the
calculus ends up with fixed values for several of the transition
variables $\TransitionVar_k$, and eventually concludes that the
transition directly from state $I_{A}$ to $F_{A}$ (variable~$x_3$)
is incompatible with
the constraint.  It is then possible to remove the $\Connected{}$
predicate using \Subsume{}, since \Propagate{} is not able to infer
further constraints, and no non-trivial applications of \Split{}
remain to be done.  After this, by further arithmetic reasoning we can
derive a solution~$l_A = 4, i = 1, n = 2$, and conclude that the root
constraint is satisfiable.  To obtain values for the string
variables~$s_1, s_2$ corresponding to the solution, one can
construct an accepting path of the automaton with each transition~$t$
taken $\Filter(t)$ times.

\iffalse

the rest of the reasoning can be
continued without any part of \Calculus{} since all values of the
existentially quantified transition variables
$\TransitionVar_1,\ldots,\TransitionVar_5$ compatible with
\cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring,const:something-before-and-after}
are represented by the flow equations introduced by $\Expand{}$.
\fi

\begin{figure}[tb]
  \centering
\begin{prooftree}
  \hypo[]{$\vdots$}
  \infer1[]{
      x_1 = 1 \land
      x_2 = l_A - 3 \land
      x_3 = 0 \land 
      x_4 = 1 \land
      x_5 = 1 \land
      n = 2 \land i = 1
    \land l_A > 3
  }
  \infer1[\Subsume]{
    \begin{array}{c}
      \Connected(A', \Filter) \land \mbox{}\\
      x_1 = 1 \land
      x_2 = l_A - 3 \land
      x_3 = 0 \land 
      x_4 = 1 \land
      x_5 = 1 \land
      n = 2 \land i = 1
    \land l_A > 3
    \end{array}
%    \begin{aligned}
%    i = \TransitionVar_5 \land 
%    n = 2 \land \\
%    0 < i < 2
%    \land l_A - i - 2 > 0 
%    \land \TransitionVar_2 = l_A - i - 2 
%    \land \TransitionVar_3 = 0 \\
%  \end{aligned}
  }
  \infer1[\EquationReasoning: $=$-simplification]{
    \begin{array}{c}
      \Connected(A', \Filter) \land\mbox{}\\
      x_1 = n - 1 \land
      x_2 = l_A - n - i \land
      x_3 = 2 - n \land 
      x_4 = n - 1 \land
      x_5 = i \land \mbox{}\\
      n = 2 \land i = 1
    \land l_A > 3
    \end{array}
%    \begin{aligned}
%    \Connected(\mathcal{A}', \Filter) \land 
%    \TransitionVar_1 = 1 \land 
%    i = \TransitionVar_5 \land 
%    n = 2 \land \\
%    0 < i < 2 
%    \land l_A - i - 2 > 0
%    \land \TransitionVar_2 = l_A - i - 2 
%    \land \TransitionVar_3 = 0
%    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $>$-reasoning]{
    \begin{array}{c}
      \Connected(A', \Filter) \land\mbox{}\\
      x_1 = n - 1 \land
      x_2 = l_A - n - i \land
      x_3 = 2 - n \land 
      x_4 = n - 1 \land
      x_5 = i \land \mbox{}\\
      n > i 
    \land l_A - i - n > 0
    \land i > 0
    \end{array}
%    \begin{aligned}
%    \Connected(\mathcal{A}', \Filter) \land
%    i = \TransitionVar_5 \land 
%    n = 2\TransitionVar_1 
%    \land n > i > 0 \\
%    \land l_A - i - n > 0 
%    \land \TransitionVar_2 = l_A - i - n 
%    \land \TransitionVar_3 = 1 - \TransitionVar_1 
%    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-simplification]{
    \begin{array}{c}
      \Connected(A', \Filter) \land\mbox{}\\
      1 = x_4 + x_3 \land x_4 = x_1 \land x_3 + x_1 = 1 \land \mbox{}\\
      l_A = x_1 + x_2 + x_3 + x_4 + x_5 \land
      i = x_5 \land 
      n = x_1 + x_3 + x_4 \land \mbox{}\\
      n > i 
    \land l_A - i - n > 0
    \land i > 0
    \end{array}
%    \begin{aligned}
%    \Connected(\mathcal{A}', \Filter)  
%    \land i = \TransitionVar_5 
%    \land n = \TransitionVar_4 + \TransitionVar_1 
%    \land n > i > 0
%    \land l_A - i - n > 0
%    \land \TransitionVar_4 = \TransitionVar_1 \\
%    \land \TransitionVar_2 = l_A - i - n 
%    \land \TransitionVar_3 = 1 - \TransitionVar_1
%    \land 1 = \TransitionVar_3 + \TransitionVar_1 
%    \land i > 0
%    \land \TransitionVar_4 = \TransitionVar_1 
%    \end{aligned}
  }
  \infer1[(Expanding $\FlowEq$)]{
    \begin{array}{c}
      \Connected(A', \Filter) \land
      \FlowEq(A', \Filter) \land \mbox{}\\
      l_A = x_1 + x_2 + x_3 + x_4 + x_5 \land
      i = x_5 \land 
      n = x_1 + x_3 + x_4 \land \mbox{}\\
      n > i 
    \land l_A - i - n > 0
    \land i > 0
    \end{array}
%    \begin{aligned}
%    i = \TransitionVar_5 
%    \land l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k 
%    \land n > i > 0
%    \land l_A - i - n > 0
%    \land n = \TransitionVar_4 + \TransitionVar_1\\
%    \land \Connected(\mathcal{A}', \Filter) 
%    \land \FlowEq(\mathcal{A}', \Filter)
%    \end{aligned}
  }
  \infer1[(Expanding sum)]{
    \begin{array}{c}
      \Connected(A', \Filter) \land
      \FlowEq(A', \Filter) \land
      \left[l_A, i, n\right] =
      \sum\limits_{\Transition \in \Transitions_{\mathcal{A}'} } \Map(\Transition) \cdot \Filter(\Transition) \land\mbox{}\\
     n > i 
    \land l_A - i - n > 0
    \land i > 0
    \end{array}
  }
  \infer1[\Expand{}]{
    \Image{}_{\mathcal{A}', \Map}(\Filter, \left[l_A, i, n\right]) 
    \land n > i 
    \land l_A - i - n > 0
    \land i > 0
    }
\end{prooftree}
\caption{A proof tree for
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $\mathcal{A}'$ of \cref{fig:aut_a}.}\label{fig:derivation:single}
\end{figure}

### 2023-07-11 multiple.tex
restructuring

>>>>

Additionally, for an
automaton~$\Automaton = \AutomatonTuple$, we allow mapping the selection function
like so: $\Filter(\Automaton) = \Tuple{\States, \InitialState, \AcceptingStates,
\Set{\Transition \in \Transitions \SuchThat \Filter(\Transition) > 0}}$, i.e.
$\Automaton$~with only the transitions for which~$\Filter$ is positive. In this
instance, the basis for the matched linear inequalities is
implicitly~$\SomeInequalities$.
